# Performing arithmetics on Hi-C matrices

```{r}
#| echo: false
source("_common.R")
```

**Aim:** This notebook illustrates how to leverage `HiCExperiment` data structure 
with `HiContacts` package to operate on Hi-C matrices. 

## Data import and basic matrix arithmetics

### Yeast data

We will start by importing a WT `.mcool` file in R and perform normalization on 
the imported matrix.

```{r yeast_norm}
library(HiCExperiment)
library(HiContactsData)
library(HiContacts)
mcf <- HiContactsData('yeast_wt', 'mcool')
pf <- HiContactsData('yeast_wt', 'pairs.gz')
cf <- CoolFile(mcf, resolution = 1000, pairsFile = pf)
wt <- import(cf)
contacts <- wt[c('II')] |> normalize(niters = 200)
contacts
cowplot::plot_grid(
    plotMatrix(contacts, use.scores = 'count', scale = 'log10'),
    plotMatrix(contacts, use.scores = 'ICE', scale = 'log10', limits = c(-4, -1))
)
```

We can do the same for a Hi-C dataset generated in a eco1 mutant. 

```{r yeast_merge}
mcf <- HiContactsData('yeast_eco1', 'mcool')
pf <- HiContactsData('yeast_eco1', 'pairs.gz')
cf <- CoolFile(mcf, resolution = 1000, pairsFile = pf)
eco1 <- import(cf)
eco1
```

Now that we have these two `HiCExperiment` objects, we can: 

1. `refocus` to change the field of view then
2. `zoom` to change the resolution then
3. `detrend` to calculate ovserved/expected fold-change and 
4. `despeckle` to remove background noise by applying a smoothing kernel

Them we can proceed to divide the contact matrix of the WT sample by the eco1 sample and plot each `HiCExperiment` matrix.

```{r yeast_divide}
wt_II <- wt |> 
    refocus('II') |> 
    zoom(2000) |> 
    detrend() |> 
    despeckle(use.scores = 'detrended')
eco1_II <- eco1 |> 
    refocus('II') |> 
    zoom(2000) |> 
    detrend() |> 
    despeckle(use.scores = 'detrended')
div_contacts <- divide(wt_II, by = eco1_II) 
div_contacts
p_div <- cowplot::plot_grid(
    plotMatrix(wt_II, use.scores = 'count', scale = 'log10'), 
    plotMatrix(eco1_II, use.scores = 'balanced', scale = 'log10', limits = c(-4, -1)), 
    plotMatrix(wt_II, compare.to = eco1_II, use.scores = 'balanced', scale = 'log10', limits = c(-4, -1)), 
    plotMatrix(div_contacts, use.scores = 'balanced.fc', scale = 'log2', limits = c(-1, 1), cmap = bwrColors())
)
p_div
```

Focusing on a narrower genomic region, it may be relevant to point out specific 
topological features, such as focal loops ("dots" in the contact matrices) or 
domain borders. We can do this by providing relevant arguments to the plotting function. 

We can also `aggregate` Hi-C maps over a list of genomic coordinates, either on-diagonal (e.g. for border aggregated plots) or off-diagonal (e.g. for loop aggregated plots).

```{r yeast_detrend}
library(rtracklayer)
library(InteractionSet)
topologicalFeatures(wt_II, 'loops') <- system.file('extdata', 'S288C-loops.bedpe', package = 'HiCExperiment') |> 
    import() |> 
    makeGInteractionsFromGRangesPairs()
topologicalFeatures(wt_II, 'borders') <- system.file('extdata', 'S288C-borders.bed', package = 'HiCExperiment') |> 
    import()
contacts <- refocus(wt_II, 'II:400000-700000') |> 
    zoom(1000) |>
    detrend() |> 
    despeckle(use.scores = 'detrended', focal.size = 2)
aggr_loops <- aggregate(contacts, targets = topologicalFeatures(wt_II, 'loops'), flankingBins = 25)
aggr_borders <- aggregate(contacts, targets = topologicalFeatures(wt_II, 'borders'), flankingBins = 15)
aggr_borders
cowplot::plot_grid(
    plotMatrix(
        contacts, 
        use.scores = 'balanced'
    ),
    plotMatrix(
        contacts, 
        use.scores = 'despeckled', 
        loops = topologicalFeatures(wt_II, 'loops'), 
        borders = topologicalFeatures(wt_II, 'borders'), 
        scale = 'linear', 
        limits = c(-1, 1), 
        cmap = bwrColors()
    ),
    plotMatrix(
        aggr_loops, 
        use.scores = 'detrended', 
        scale = 'linear', 
        limits = c(-1, 1), 
        cmap = rainbowColors()
    ),
    plotMatrix(
        aggr_borders, 
        use.scores = 'detrended', 
        scale = 'linear', 
        limits = c(-0.5, 0.5), 
        cmap = bwrColors()
    )
)
```

### Playing with mammalian data

The same operations (i.e. detrending, despeckling, ,,,) are also possible on larger genomes. 
On top of that, it is sometimes useful to `autocorrelate` the Hi-C matrix to highlight
the compartments "plaid" pattern.

Here, the `maxDistance` argument can be used to plot triangular horizontal matrices rather 
than square matrices. 

```{r NPCs, eval = FALSE}
library(fourDNData)
library(cowplot)
## G0G1 neural progenitors cells  (FACS sorted, Sox1-GFP+), Bonev et al. 2017
npcs_mcool <- fourDNData('4DNESJ9SIAV5', type = 'mcool')
npcs_large <- import(npcs_mcool, focus = 'chr3:80000000-120000000', resolution = 250000) |> 
    autocorrelate()
npcs_narrow <- import(npcs_mcool, focus = 'chr3:100800000-102400000', resolution = 5000) |> 
    detrend() |> 
    despeckle(use.scores = 'detrended')
p_mammals_1 <- plot_grid(
    plotMatrix(npcs_large, use.scores = 'balanced', scale = 'log10', limits = c(-4, -1), maxDistance = 30000000), 
    plotMatrix(npcs_large, use.scores = 'autocorrelated', scale = 'linear', limits = c(-1, 1), cmap = bgrColors(), maxDistance = 30000000), 
    plotMatrix(npcs_narrow, use.scores = 'balanced', scale = 'log10', limits = c(-4, -2), maxDistance = 1200000), 
    plotMatrix(npcs_narrow, use.scores = 'expected', scale = 'log10', limits = c(-4, -2), maxDistance = 1200000), 
    plotMatrix(npcs_narrow, use.scores = 'detrended', scale = 'linear', limits = c(-1, 1), maxDistance = 1200000, cmap = rainbowColors()), 
    plotMatrix(npcs_narrow, use.scores = 'despeckled', scale = 'linear', limits = c(-2, 2), maxDistance = 1200000, cmap = rainbowColors()), 
    ncol = 1
)
p_mammals_1
```

Finally, `subsample`ing interactions from a Hi-C experiment can easily be 
performed, and this is done proportionally with the distance-dependent 
interaction frequency. Furthermore, if a Hi-C experiment has been binned to a small resolution, 
it can be `coarsen`ed to any larger resolution. 

```{r subsample, eval = FALSE}
sub_npcs_0.5 <- subsample(npcs_narrow, 0.5)
sub_npcs_0.1 <- subsample(npcs_narrow, 0.1) |> 
    detrend()
sub_npcs_0.1_rebinned <- coarsen(sub_npcs_0.1, 10000) |> 
    detrend() 
p_mammals_2 <- cowplot::plot_grid(
    plotMatrix(npcs_narrow, use.scores = 'balanced', limits = c(-4, -2), maxDistance = 1200000), 
    plotMatrix(sub_npcs_0.5, use.scores = 'balanced', limits = c(-4, -2), maxDistance = 1200000), 
    plotMatrix(sub_npcs_0.1, use.scores = 'balanced', limits = c(-4, -2), maxDistance = 1200000), 
    plotMatrix(sub_npcs_0.1, use.scores = 'detrended', scale = 'linear', limits = c(-1, 1), maxDistance = 1200000, cmap = rainbowColors()), 
    plotMatrix(sub_npcs_0.1_rebinned, use.scores = 'balanced', maxDistance = 1200000), 
    plotMatrix(sub_npcs_0.1_rebinned, use.scores = 'detrended', scale = 'linear', limits = c(-1, 1), maxDistance = 1200000, cmap = rainbowColors()), 
    ncol = 1
)
p_mammals_2
```

