# Representation of genomic data in R

```{r}
#| echo: false
source("_common.R")
```

## `GRanges` class

## `GInteractions` class

## `ContactFile` class

Contact matrices can be in three different formats (see previous chapter): 

- `.(m)cool` matrix (multi-scores, multi-resolution, indexed)
- `.hic` matrix (multi-scores, multi-resolution, indexed)
- HiC-pro derived matrix (single-score, single-resolution, non-indexed)

Example contact matrix files can be obtained using `HiContactsData` function. 

```{r}
library(HiContactsData)
coolf <- HiContactsData('yeast_wt', 'mcool')
hicf <- HiContactsData('yeast_wt', 'hic')
hicpromatrixf <- HiContactsData('yeast_wt', 'hicpro_matrix')
hicproregionsf <- HiContactsData('yeast_wt', 'hicpro_bed')
```

A connection to each of these files can then be created in R using the matching function. 

```{r}
library(HiCExperiment)
CoolFile(coolf)
HicFile(hicf)
HicproFile(hicpromatrixf, hicproregionsf)
```

## `HiCExperiment` class

```{r yeast_norm}
hic <- contacts_yeast(full = TRUE)
hic
```
