# Welcome {-}

```{r}
#| echo: false
source("_common.R")
.gh_url <- file.path('https://github.com', rmarkdown::metadata[['github-repo']])
```

`r rebook::openingDetails(Copyright='J. Serizay', Source=.gh_url)`

This is the landing page of the **"Orchestrating Hi-C analysis with Bioconductor"** 
book. **The primary aim of this book is to introduce the `R` user to Hi-C analysis**. It 
starts with key concepts important for the analysis of chromatin conformation capture and then 
presents `Bioconductor` tools that can be leveraged to process, analyse, explore and visualize 
Hi-C data. 

::: {.callout-caution}
This project is currently under active development. Several sections of this book may not be accurate, as the packages introduced in this book are still maturing. 
:::

# Table of contents {-}

This book is divided in three parts:  

**Part I: Introduction to Hi-C analysis**

- [Chapter 1: General principles and Hi-C data pre-processing](principles.qmd)
- [Chapter 2: The different R classes implemented to analyse Hi-C](data-representation.qmd)
- [Chapter 3: Manipulating Hi-C data in R](parsing.qmd)
- [Chapter 4: Hi-C data visualization](visualization.qmd)


**Part II: In-depth Hi-C analysis**

- [Chapter 5: Matrix-centric analysis](matrix-centric.qmd)
- [Chapter 6: Interactions-centric analysis](interactions-centric.qmd)
- [Chapter 7: Finding topological features from a Hi-C contact matrix](topological-features.qmd)


**Part III: Hi-C analysis workflows**

- [Chapter 8: Distance-dependent interactions across yeast mutants](workflow-yeast.qmd)
- [Chapter 9: Chromosome compartment cohesion upon mitosis entry](workflow-chicken.qmd)
- [Chapter 10: Hi-C interoperability notebook](interoperability.qmd)
- [Chapter 11: Access to public Hi-C data portals](disseminating.qmd)

# Requirements {-}

This books aims to demonstrate how to pre-process, parse and investigate Hi-C data in `R`. 
For this reason, a significant portion of this book consists of executable R code chunks. 
To be able to reproduce the examples demonstrated in this book and go further in the analysis of 
*your **real** datasets*, you will need to rely on several dependencies. 

- The most important dependency is `R` itself! Make sure you have `R >= 4.3`. You can check that by typing `version` in an R console. If you do not have `R >= 4.3` installed, we highly recommend to update your `R` version, as most extra dependencies 
will require `R >= 4.3`. 

- The following `R` code will set up most of the extra dependencies: 

```{r eval = FALSE}
BiocManager::install("HiCExperiment", ask = FALSE)
BiocManager::install("HiCool", ask = FALSE)
BiocManager::install("HiContacts", ask = FALSE)
BiocManager::install("HiContactsData", ask = FALSE)
BiocManager::install("fourDNData", ask = FALSE)
BiocManager::install("DNAZooData", ask = FALSE)
```

For developers or advanced R users, there are two alternatives: 

- You can directly install `devel` versions of these packages along with their dependencies: 

```{r eval = FALSE}
install.packages("pak", repos = "https://r-lib.github.io/p/pak/devel/")
pak::pkg_install("js2264/HiCExperiment", ask = FALSE, dependencies = c("Depends", "Imports", "Suggests"))
pak::pkg_install("js2264/HiCool", ask = FALSE, dependencies = c("Depends", "Imports", "Suggests"))
pak::pkg_install("js2264/HiContacts", ask = FALSE, dependencies = c("Depends", "Imports", "Suggests"))
pak::pkg_install("js2264/HiContactsData", ask = FALSE, dependencies = c("Depends", "Imports", "Suggests"))
pak::pkg_install("js2264/fourDNData", ask = FALSE, dependencies = c("Depends", "Imports", "Suggests"))
pak::pkg_install("js2264/DNAZooData", ask = FALSE, dependencies = c("Depends", "Imports", "Suggests"))
```

- Alternatively, if you have `docker` installed, the easiest approach would be to run the following command in a `shell` terminal:

```{sh eval = FALSE}
docker run -it ghcr.io/js2264/ohca:latest R
```

This will fetch a `docker` image with the latest development versions of the 
aformentioned packages pre-installed, and initate an interactive R session. 

# Reproducibility {-}

## Building book {-}

The OHCA book has been rendered with R thanks to a number of packages, including but not only: 

- `devtools`
- `quarto`
- `rebook`

To build this book locally, you can run: 

```{sh eval = FALSE, filename="bash"}
git clone git@github.com:js2264/OHCA.git && cd OHCA
quarto render
```

::: {.callout-warning icon='false'}
All dependencies listed above will be required!
:::

In addition, the actual book rendering is done by GitHub Actions, and the 
rendered static website is hosted by GitHub Pages. 

## Session info {-}

```{r echo = FALSE, warnings = FALSE, messages = FALSE}
library("HiCExperiment")
library("HiCool")
library("HiContacts")
library("HiContactsData")
library("fourDNData")
library("DNAZooData")
```

```{r echo = FALSE}
sessioninfo::session_info()
```
