--- 
title: "Orchestrating Hi-C analysis with Bioconductor"
documentclass: book
biblio-style: apalike
link-citations: yes
github-repo: js2264/OHCA
---

```{r, echo=FALSE}
.gh_url <- file.path('https://github.com', rmarkdown::metadata[['github-repo']])
```

---
date: "`r rebook::openingDetails(Copyright='J. Serizay', Source=.gh_url)`"
url: "`r .gh_url`"
---

# Introduction {-}

This site is the web-based version of the book entitled "Orchestrating Hi-C analysis with Bioconductor".  

In this book, we will cover fundamentals topis: 

- What Hi-C is experimentally and computationally 
- The different `R` classes implemented to analyse Hi-C
- Parsing and manipulating Hi-C data in R
- Hi-C data visualization

We will then focus on more Hi-C-focused questions: 

- Performing "arithmetics" on Hi-C matrices
- Computing Hi-C-specific metrics
- Finding topological features from a Hi-C contact matrix

The following advanced topics will also be introduced: 

- Hi-C interoperability in R/Bioconductor

```{r include=FALSE}
knitr::opts_chunk$set(
    cache = TRUE, 
    layout = 'l-page', 
    echo = TRUE, 
    warning = FALSE, 
    error = FALSE, 
    message = FALSE
)
```
