---
reference-section-title: References
bibliography: bibliography.bib
---

# Interoperability: using HiCExperiment with other R packages

::: {.callout-note}
## Aims

This notebook illustrates how to use a range of popular Hi-C—related `R` packages with 
`HiCExperiment` objects. Conversion to the following packages is 
illustrated here: 

- `hicrep`
- `multiHiCcompare`
- `TopDom`
- `GOTHiC`
<!-- - `HiCDOC` -->

:::

```{r}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false
source("_common.R")
library(ggplot2)
library(purrr)
library(GenomicRanges)
library(InteractionSet)
library(HiCExperiment)
library(HiContactsData)
library(hicrep)
library(multiHiCcompare)
library(dplyr)
library(tidyr)

coolf_wt <- HiContactsData('yeast_wt', 'mcool')
coolf_eco1 <- HiContactsData('yeast_eco1', 'mcool')
```

## HiCrep

`hicrep` is a popular package to compute **stratum-adjusted correlations** between Hi-C datasets (@Yang2017Nov). 
"Stratum" refers to the distance from the main diagonal: with increase distance from the main diagonal, 
interactions of the DNA polymer are bound to decrease. `hicrep` computes a "per-stratum" correlation score and 
computes a weighted average correlation for entire chromosomes. 

::: {.callout-tip}
### Installing `hicrep`

`hicrep` package has been available from Bioconductor for many years but has 
been withdrawn from their repositories at some point. You can always install 
`hicrep` directly from its GitHub repository as follows: 

```{r}
#| eval: false
remotes::install_github('TaoYang-dev/hicrep')
```

:::

In order to use `hicrep`, we first need to create two `HiCExperiment` objects. 

```{r eval = FALSE}
library(HiCExperiment)
library(HiContactsData)

# ---- This downloads example `.mcool` and `.pairs` files and caches them locally 
coolf_wt <- HiContactsData('yeast_wt', 'mcool')
coolf_eco1 <- HiContactsData('yeast_eco1', 'mcool')
```

```{r}
hic_wt <- import(coolf_wt, format = 'cool')
hic_eco1 <- import(coolf_eco1, format = 'cool')
```

We can now run the main `get.scc` function from `hicrep`. The documentation 
for this function is available from the console by typing `?hicrep::get.scc`. 
More information is also available from the [GitHub page](https://github.com/TaoYang-dev/hicrep#compute-reproducibility-score-using-hicrep). It informs the end user that 
the input for this function should be two intra-chromosomal Hi-C **raw count** matrices 
in square (optionally sparse) format. 

```{r}
hic_wt

as.matrix(hic_wt["IV"], use.scores = 'count')[1:10, 1:10]

library(hicrep)
scc <- get.scc(
    as.matrix(hic_wt["IV"], use.scores = 'count'), 
    as.matrix(hic_eco1["IV"], use.scores = 'count'), 
    resol = 1000, h = 25, lbr = 5000, ubr = 50000
)
scc

scc$scc
```

## multiHiCcompare

The `multiHiCcompare` package provides functions for joint 
normalization and difference detection in multiple Hi-C datasets (@Stansfield2019Sep). 
According to its excerpt, to perform differential interaction analysis, 
it requires a `list` of **raw counts** for different 
samples/replicates, stored in **data frames with four columns** (`chr`, `start1`, `start2`, `count`).  
Manipulate a `HiCExperiment` object to coerce it into such structure is straightforward. 

```{r}
library(dplyr)
library(tidyr)
hics <- list(
    "wt" = import(coolf_wt, format = 'cool'),
    "eco1" = import(coolf_eco1, format = 'cool')
)
hics_list <- map(hics, ~ .x['XI'] |> 
    as.data.frame() |>
    mutate(chr = 1) |> 
    relocate(chr) |>
    select(chr, start1, start2, count)
)
head(hics_list[[1]])
```

Once this list is generated, the classical `multiHiCcompare` workflow can be applied: 
first run `make_hicexp()`, followed by `cyclic_loess()`, then `hic_exactTest()` and finally `results()`: 

```{r}
#| warning: false
DI <- hics_list |> 
    make_hicexp(
        data_list = hics_list, 
        groups = factor(c(1, 2))
    ) |> 
    cyclic_loess() |> 
    hic_exactTest() |> 
    results()
DI
```

## TopDom

The `TopDom` method is widely used to annotate topological domains in genomes from Hi-C 
data (@Shin2016Apr). The `TopDom` package was created to implement this method 
in `R` (@BibEntry2021May).

Unfortunately, the format of the input to `TopDom` is rather tricky (see `?TopDom::readHiC`). 
The following chunk of code shows how to coerce a `HiCExperiment` object into a 
`TopDom`-compatible object. 


```{r}
library(TopDom)
hic <- import(coolf_wt, format = 'cool')
HiCExperiment2TopDom <- function(hic, chr) {
    data <- list()
    cm <- as(hic[chr], 'ContactMatrix')
    data$counts <- as.matrix(cm) |> base::as.matrix()
    data$counts[is.na(data$counts)] <- 0
    data$bins <- regions(cm) |> 
        as.data.frame() |> 
        select(seqnames, start, end) |>
        mutate(seqnames = as.character(seqnames)) |>
        mutate(id = 1:n(), start = start - 1) |> 
        relocate(id) |> 
        dplyr::rename(chr = seqnames, from.coord = start, to.coord = end)
    class(data) <- 'TopDomData'
    return(data)
}
hic_topdom <- HiCExperiment2TopDom(hic, "II")
hic_topdom
```

Now that we have coerced a `HiCExperiment` object into a `TopDom`-compatible 
object, we can use the main `TopDom` function to annotate topological domains. 

```{r}
domains <- TopDom::TopDom(hic_topdom, window.size = 5)
domains
```

The resulting `domains` object can be used to extract annotated domains, 
store them in `topologicalFeatures` of the original `HiCExperiment`, and 
optionally write a `bed` file to export them in text. 

```{r}
topologicalFeatures(hic, 'domain') <- domains$bed |> 
    mutate(chromStart = chromStart + 1) |> 
    filter(name == 'domain') |> 
    makeGRangesFromDataFrame()
topologicalFeatures(hic, 'domain')

rtracklayer::export(topologicalFeatures(hic, 'domain'), 'hic_domains.bed')
```

## GOTHiC

`GOTHiC` relies on a cumulative binomial test to detect interactions between distal genomic loci that have significantly more reads than expected by chance in Hi-C experiments (@Mifsud2017Apr). 

::: {.callout-important icon='true'}
### Using the `GOTHiC` function

Unfortunately, the main `GOTHiC` function 
require two `.bam` files as input. These files are often deleted due to 
their larger size, while the filtered `pairs` file itself is retained. 

Moreover, the internal nuts and bolts of the main `GOTHiC` function perform 
several operations that are not required in modern workflows: 

1. [Filtering pairs from same restriction fragment](https://code.bioconductor.org/browse/GOTHiC/blob/RELEASE_3_17/R/GOTHiC.R#L814); this step is now usually taken care of automatically, e.g. with `HiCool` Hi-C processing package. 
2. [Filtering short-range pairs](https://code.bioconductor.org/browse/GOTHiC/blob/RELEASE_3_17/R/GOTHiC.R#L826); the `GOTHiC` package hard-codes a 10kb lower threshold for minimum pair distance. More advanced optimized filtering approaches have been implemented since then, to circumvent the need for such hard-coded threshold.
3. [Binning pairs](https://code.bioconductor.org/browse/GOTHiC/blob/RELEASE_3_17/R/GOTHiC.R#L834); this step 
is also already taken care of, when working with Hi-C matrices in modern formats, e.g. with `.(m)cool` files. 

:::

Based on these facts, we can simplify the binomial test function provided by 
`GOTHiC` so that it can directly used binned interactions imported as a 
`HiCExperiment` object in `R`. 

```{r}
#| code-fold: true
#| code-summary: "Show the code for `GOTHiC_binomial` function"
GOTHiC_binomial <- function(x) {

    if (length(trans(x)) != 0) stop("Only `cis` interactions can be used here.")
    ints <- interactions(x) |>
        as.data.frame() |> 
        select(seqnames1, start1, seqnames2, start2, count) |>
        dplyr::rename(chr1 = seqnames1, locus1 = start1, chr2 = seqnames2, locus2 = start2, frequencies = count) |>
        mutate(locus1 = locus1 - 1, locus2 = locus2 - 1) |>
        mutate(int1 = paste0(chr1, '_', locus1), int2 = paste0(chr2, '_', locus2))
    
    numberOfReadPairs <- sum(ints$frequencies)
    all_bins <- unique(c(unique(ints$int1), unique(ints$int2)))
    all_bins <- sort(all_bins)
    upperhalfBinNumber <- (length(all_bins)^2 - length(all_bins))/2

    cov <- ints |> 
        group_by(int1) |> 
        tally(frequencies) |> 
        full_join(ints |> 
            group_by(int2) |> 
            tally(frequencies), 
            by = c('int1' = 'int2')
        ) |> 
        rowwise() |> 
        mutate(coverage = sum(n.x, n.y, na.rm = TRUE)) |> 
        ungroup() |>
        mutate(relative_coverage = coverage/sum(coverage))
    
    results <- mutate(ints,
        cov1 = left_join(ints, select(cov, int1, relative_coverage), by = c('int1' = 'int1'))$relative_coverage, 
        cov2 = left_join(ints, select(cov, int1, relative_coverage), by = c('int2' = 'int1'))$relative_coverage,
        probability = cov1 * cov2 * 2 * 1/(1 - sum(cov$relative_coverage^2)),
        predicted = probability * numberOfReadPairs
    ) |> 
    rowwise() |>
    mutate(
        pvalue = binom.test(
            frequencies, 
            numberOfReadPairs, 
            probability,
            alternative = "greater"
        )$p.value
    ) |> 
    ungroup() |> 
    mutate(
        logFoldChange = log2(frequencies / predicted), 
        qvalue = stats::p.adjust(pvalue, method = "BH", n = upperhalfBinNumber)
    )

    scores(x, "probability") <- results$probability
    scores(x, "predicted") <- results$predicted
    scores(x, "pvalue") <- results$pvalue
    scores(x, "qvalue") <- results$qvalue
    scores(x, "logFoldChange") <- results$logFoldChange

    return(x)

} 
```

```{r}
res <- GOTHiC_binomial(hic["II"])
res

interactions(res)
```

<!-- ## HiCDOC -->

## References {-}

::: {#refs}
:::

## Session info {-}

```{r echo = FALSE}
sessioninfo::session_info()
```
