# Hi-C data visualization

```{r}
#| echo: false
source("_common.R")
```

## Visualizing Hi-C contact maps

Visualizing Hi-C contact maps is often a necessary step in exploratory data 
analysis. A Hi-C contact map is usually displayed as a heatmap, in which: 

- Each axis represents a section of the genome of interest (either a segment of a chromosome, or 
several chromosomes, ...).
- The color code aims to represent "interaction frequency", which can be expressed 
in "raw" counts or normalized (balanced). 
- Other metrics can also be displayed in Hi-C heatmaps, e.g. ratios of interaction 
frequency between two Hi-C experiments, p-values of differential interaction analysis, ... 
- Axes are often identical, representing interactions constrained within a single genomic
window, a.k.a. **on-diagonal** matrices. 
- However, axes *can* be different: this is the case when **off-diagonal** matrices 
are displayed. 

### Single map

### Horizontal map

### Side-by-side maps

### Aggregated Hi-C map

## Hi-C maps customization options

### Choosing scores

### Choosing color map

### Choosing scale

## References {-}

::: {#refs}
:::

## Session info {-}

```{r echo = FALSE}
sessioninfo::session_info()
```
