# Interoperability with `python`: `cooltools` {#sec-python-cooltools}

::: {.callout-note collapse="true"}

## Pre-loading packages and objects 📦 {-}

```{r}
## `reticulate` is attached first, so that `import()` refers to
## `HiCExperiment`'s rather than to `reticulate`'s
library(reticulate)
## `python` packages are imported before other `R` packages are attached:
## these can load the system's graphics libraries (e.g. `harfbuzz`), and
## `matplotlib` from `conda` then fails to import if it needs a more recent
## version of them
BiocBook::setup_python()
invisible(reticulate::import("matplotlib"))
library(dplyr)
library(ggplot2)
library(patchwork)
library(GenomicRanges)
library(HiCExperiment)
library(HiContactsData)
library(HiContacts)
library(BiocParallel)
```

:::

::: {.callout-note}

## Aims

This chapter shows how `python` tools dedicated to Hi-C analysis can be used
side by side with `HiCExperiment` and `HiContacts`, within a single analysis:

- Running `cooltools` (@Open2C_2024) on a `.mcool` file fetched in `R`;
- Passing objects between `R` and `python`, in both directions;
- Comparing the insulation scores and A/B compartments computed by `HiContacts`
  and by `cooltools` from the same data.

:::

## Why `cooltools`?

`cooltools` is the `python` toolkit developed by the Open2C community to
analyze contact matrices stored in `.cool`/`.mcool` files, built on top of
`cooler` (@Abdennur_2019) and `bioframe` (@Open2C_2024_bioframe). It is the
reference implementation of many analyses covered in this book, and the 4DN
consortium uses it to compute the insulation and compartment tracks it
distributes (see the [chapter on public Hi-C data portals](disseminating.qmd)).

Running it next to `HiContacts` is useful to compare results, to reproduce
published analyses, or to use features only available in one of them. The
`python` code of this chapter is executed every time the book is built,
including when the *Bioconductor Build System* rebuilds it.

## Setting up `python`

`python` chunks of this book are executed through `reticulate`, in a `python`
session living alongside the `R` session. `R` objects are available from
`python` as `r.<object>`, and `python` objects from `R` as `py$<object>`.

::: {.callout-tip icon='true'}

## Installing `python` packages for this book

The `python` packages used in this chapter are declared, with pinned versions,
in the `conda` environment file of the book (`inst/requirements.yml`):

```{r}
#| echo: false
#| comment: ""
cat(readLines("../requirements.yml"), sep = "\n")
```

`BiocBook::setup_python()`, called at the top of this chapter, creates this
environment if needed (fetching a standalone `micromamba` if it finds none on
the machine) and activates it with `reticulate`. In the book's
`Docker` image, the environment is already installed and activated.

:::

## Fetching data in `R`, reading it in `python`

This chapter uses the micro-C dataset (@Krietenstein_2020) already analyzed in
the [chapter on topological features](topological-features.qmd). It contains
intra-chromosomal interactions within `chr17`, binned at `5000`, `100000` and
`250000` bp. The `.mcool` file is fetched from `R`...

```{r}
mcool <- unname(HiContactsData('microC', 'mcool'))
mcool
```

... and opened in `python` with `cooler`, using the file path defined in `R`:

```{python}
import cooler
clr = cooler.Cooler(f"{r.mcool}::/resolutions/5000")
clr.info
```

## Insulation and domain boundaries

`cooltools.insulation()` computes the diamond insulation score (@Crane_2015) and
calls domain boundaries. To match the analysis of the
[topological features chapter](topological-features.qmd), it is run on the
long arm of `chr17` (from 60 Mb), at 5 kb resolution, with a 100 kb window.
`bioframe` defines this genomic region as a "view":

```{python}
import bioframe
import cooltools
view_chr17q = bioframe.make_viewframe([("chr17", 60_000_000, 83_257_441, "chr17q")])
insulation = cooltools.insulation(clr, [100_000], view_df = view_chr17q, verbose = False)
insulation[["chrom", "start", "end", "log2_insulation_score_100000", "is_boundary_100000"]].dropna()
```

`cooltools` called `r sum(py$insulation$is_boundary_100000)` domain boundaries 
on this arm. 

## A/B compartments: from `R` to `python` and back

`cooltools.eigs_cis()` computes the eigenvectors of the contact matrix, and
orients them with a "phasing track" (such as the GC content of each genomic
bin), so that positive values correspond to the A compartment.

The 250 kb bins of the contact matrix are listed in `python`...

```{python}
clr_250kb = cooler.Cooler(f"{r.mcool}::/resolutions/250000")
bins = clr_250kb.bins()[:][["chrom", "start", "end"]]
```

... their GC content is computed in `R`, from the `BSgenome` reference
sequence used throughout this book...

```{r}
hg38 <- BSgenome.Hsapiens.UCSC.hg38::BSgenome.Hsapiens.UCSC.hg38
bins <- makeGRangesFromDataFrame(py$bins, starts.in.df.are.0based = TRUE)
GC <- Biostrings::letterFrequency(
    Biostrings::getSeq(hg38, bins), letters = "GC", as.prob = TRUE
)[, 1]
gc_track <- data.frame(py$bins, GC = GC)
head(gc_track)
```

... and the GC track is passed back to `python` to phase the eigenvectors:

```{python}
#| warning: false
view_chr17 = bioframe.make_viewframe([("chr17", 0, 83_257_441, "chr17")])
eigenvalues, eigenvectors = cooltools.eigs_cis(
    clr_250kb, phasing_track = r.gc_track, view_df = view_chr17, n_eigs = 3
)
eigenvectors.dropna().head()
```

## Do `R` and `python` agree?

The same analyses are run with `HiContacts`, exactly as in the
[chapter on topological features](topological-features.qmd):

```{r}
microC <- import(CoolFile(mcool), resolution = 250000)
microC_compts <- getCompartments(microC, genome = hg38)
hic <- zoom(microC, 5000) |>
    refocus('chr17:60000001-83257441') |>
    getDiamondInsulation(window_size = 100000, BPPARAM = SerialParam(progressbar = FALSE)) |>
    getBorders()
```

The `python` results are retrieved in `R` as `data.frame`s, and matched to the
`HiContacts` results bin by bin (`python` coordinates are 0-based, `R`
coordinates are 1-based):

```{r}
insulation <- tibble(
    start = start(metadata(hic)$insulation) - 1,
    HiContacts = metadata(hic)$insulation$insulation
) |> inner_join(
    tibble(start = py$insulation$start, cooltools = py$insulation$log2_insulation_score_100000),
    by = "start"
)
compartments <- tibble(
    start = start(metadata(microC_compts)$eigens) - 1,
    HiContacts = metadata(microC_compts)$eigens$eigen
) |> inner_join(
    tibble(start = py$eigenvectors$start, cooltools = py$eigenvectors$E1),
    by = "start"
)
cors <- c(
    insulation = cor(insulation$HiContacts, insulation$cooltools, use = "complete.obs"),
    compartments = cor(compartments$HiContacts, compartments$cooltools, use = "complete.obs")
)
cors
```

```{r}
#| label: fig-r-vs-python
#| fig-cap: "Insulation scores (top, `chr17` long arm, 5 kb) and A/B compartment eigenvectors (bottom, `chr17`, 250 kb) computed by `HiContacts` (`R`) and by `cooltools` (`python`) from the same micro-C contact matrix."
#| fig-width: 9
#| fig-height: 6
tracks <- function(df, ylab) {
    tidyr::pivot_longer(df, c(HiContacts, cooltools), names_to = "tool") |>
        ggplot(aes(x = start / 1e6, y = value, colour = tool)) +
        geom_line(linewidth = 0.3, na.rm = TRUE) +
        scale_colour_manual(values = c(HiContacts = "#0484a9", cooltools = "#e2711d")) +
        labs(x = "chr17 (Mb)", y = ylab, colour = NULL) +
        theme_bw() +
        theme(legend.position = "top")
}
scatter <- function(df, r) {
    ggplot(df, aes(x = HiContacts, y = cooltools)) +
        geom_point(size = 0.4, alpha = 0.3, na.rm = TRUE) +
        annotate("text", x = -Inf, y = Inf, hjust = -0.2, vjust = 1.5,
            label = sprintf("r = %.3f", r)) +
        labs(x = "HiContacts (R)", y = "cooltools (python)") +
        theme_bw()
}
wrap_plots(
    tracks(insulation, "Insulation score"), scatter(insulation, cors[["insulation"]]),
    tracks(compartments, "Eigenvector"), scatter(compartments, cors[["compartments"]]),
    ncol = 2, widths = c(3, 1), guides = "collect"
) & theme(legend.position = "top")
```

Both tools agree closely (@fig-r-vs-python). The compartment eigenvectors are 
almost identical (Pearson correlation of `r round(cors[["compartments"]], 3)`), 
and so are their signs, both being phased with the same GC track. Insulation 
scores are highly correlated as well (`r round(cors[["insulation"]], 3)`): the 
two implementations differ in how they normalize the score (by its median in 
`cooltools`, by its mean in `HiContacts`) and handle bins filtered out by the 
balancing, not in what they measure. 

Domain boundaries, on the other hand, are called from these scores with 
different procedures. Both keep the local minima of the insulation score that 
are prominent enough, but `HiContacts::getBorders()` measures each minimum 
against the next local maximum and uses a fixed threshold (`0.2` by default), 
while `cooltools.insulation()` uses the topographic prominence of each minimum 
and sets its threshold from their distribution (Li's method, by default). The 
two tools therefore call different sets of boundaries from very similar 
scores: 

```{r}
borders_HiContacts <- topologicalFeatures(hic, "borders")
borders_cooltools <- py$insulation |> 
    filter(is_boundary_100000) |> 
    makeGRangesFromDataFrame(starts.in.df.are.0based = TRUE)
c(
    HiContacts = length(borders_HiContacts), 
    cooltools = length(borders_cooltools), 
    `HiContacts borders within 10 kb of a cooltools boundary` = 
        sum(overlapsAny(borders_HiContacts, borders_cooltools, maxgap = 10000))
)
```

Comparing boundaries between studies, or between tools, thus requires using the 
same calling procedure on both sides, rather than comparing lists of boundaries 
called by different tools. 

# Session info {-}

::: {.callout-note collapse="true"}

## Click to expand 👇 {-}

```{r "session info"}
#| cache: false
sessioninfo::session_info(include_base = TRUE)
```

:::

# References {-}
