# Hi-C data visualization

```{r}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false
library(ggplot2)
library(GenomicRanges)
library(InteractionSet)
library(HiCExperiment)
library(HiContactsData)
library(HiContacts)
library(rtracklayer)
```

::: {.callout-note}

## Aims
This chapter focuses on the various visualization tools offered by `HiContacts`
to plot `HiCExperiment` contact matrices in R. 

:::

::: {.callout-tip collapse="true"}

## Generating the example `hic` object 👇

To demonstrate how to visualize a `HiCExperiment` contact matrix, we will create 
an `HiCExperiment` object from an example `.cool` file provided 
in the `HiContactsData` package. 

```{r}
library(HiCExperiment)
library(HiContactsData)

# ---- This downloads an example `.mcool` file and caches it locally 
coolf <- HiContactsData('yeast_wt', 'mcool')

# ---- This creates a connection to the disk-stored `.mcool` file
cf <- CoolFile(coolf)
cf

# ---- This imports contacts from the chromosome `V` at resolution `2000`
hic <- import(cf, focus = 'V', resolution = 2000)
```

```{r}
hic
```

:::

## Visualizing Hi-C contact maps

Visualizing Hi-C contact maps is often a necessary step in exploratory data 
analysis. A Hi-C contact map is usually displayed as a heatmap, in which: 

- Each axis represents a section of the genome of interest (either a segment of a chromosome, or several chromosomes, ...).
- The color code aims to represent "interaction frequency", which can be expressed 
in "raw" counts or normalized (balanced). 
- Other metrics can also be displayed in Hi-C heatmaps, e.g. ratios of interaction 
frequency between two Hi-C experiments, p-values of differential interaction analysis, ... 
- Axes are often identical, representing interactions constrained within a single genomic
window, a.k.a. **on-diagonal** matrices. 
- However, axes *can* be different: this is the case when **off-diagonal** matrices 
are displayed. 

### Single map

Simple visualization of disk-stored Hi-C contact matrices can be done by: 

1. Importing the interactions over the genomic location of interest into a `HiCExperiment` object;
2. Using `plotMatrix` function (provided by `HiContacts`) to generate a plot.

```{r}
library(HiContacts)
plotMatrix(hic)
```

::: {.callout-note}

#### Note
A caption summarizing the plotting parameters is added below the heatmap. This 
can be removed with `caption = FALSE`.

:::

### Horizontal map

Hi-C maps are sometimes visualized in a "horizontal" style, where 
a square **on-diagonal** heatmap is tilted by 45˚ and truncated to only show 
interactions up to a certain distance from the main diagonal. 

When a `maxDistance` argument is provided to `plotMatrix`, it automatically
generates a horizontal-style heatmap. 

```{r}
plotMatrix(hic, maxDistance = 200000)
```

### Side-by-side maps

Sometimes, one may want to visually plot 2 Hi-C samples side by side to compare
the interaction landscapes over the same genomic locus. This can be done by 
adding a second `HiCExperiment` (imported with the same 
`focus`) with the `compare.to` argument. 

Here, we are importing a second `.mcool` file corresponding to a Hi-C experiment 
performed in a *eco1* yeast mutant: 

```{r}
hic2 <- import(
    CoolFile(HiContactsData('yeast_eco1', 'mcool')), 
    focus = 'V', 
    resolution = 2000
)
```

We then plot the 2 matrices side by side. The first will be displayed in the 
top right corner and the second (provided with `compare.to`) will be in the
bottom left corner. 

```{r}
plotMatrix(hic, compare.to = hic2)
```

### Plotting multiple chromosomes

Interactions from multiple chromosomes can be visualized in a Hi-C heatmap. 
One needs to (1) first parse the entire contact matrix in `R`, (2) then subset
interactions over chromosomes of interest with `[` and (3) use `plotMatrix` to 
generate the multi-chromosome plot. 

```{r}
full_hic <- import(cf, resolution = 4000)
plotMatrix(full_hic)
hic_subset <- full_hic[c("II", "III", "IV")]
plotMatrix(hic_subset)
```

## Hi-C maps customization options

A number of customization options are available for the `plotMatrix` function. 
The next subsections focus on how to: 

- Pick the `scores` of interest to represent in a Hi-C heatmap;
- Change the numeric scale and boundaries;
- Change the color map;
- Extra customization options

### Choosing scores

By default, `plotMatrix` will attempt to plot `balanced` (coverage normalized) 
Hi-C matrices. However, extra scores may be associated with interactions in a 
`HiCExperiment` object (more on this in the [next chapter](matrix-centric.qmd))

For instance, we can plot the `count` scores, which are un-normalized raw 
contact counts directly obtained when binning a `.pairs` file: 

```{r}
plotMatrix(hic, use.scores = 'count')
```

### Choosing scale

The color scale is automatically adjusted to range from the minimum to the 
maximum `scores` of the `HiCExperiment` being plotted. This can be adjusted 
using the `limits` argument. 

```{r}
plotMatrix(hic, limits = c(-3.5, -1))
```

### Choosing color map

`?HiContacts::palettes` returns a list of available color maps to use 
with `plotMatrix`. Any custom color map can also be used by manually 
specifying a vector of colors. 

```{r}
# ----- `afmhotr` color map is shipped in the `HiContacts` package
afmhotrColors() 
plotMatrix(
    hic, 
    use.scores = 'balanced',
    limits = c(-4, -1),
    cmap = afmhotrColors()
)
```

## Advanced visualization 

### Overlaying topological features 

Topological features (e.g. chromatin loops, domain borders, A/B compartments, 
e.g. ...) are often displayed over a Hi-C heatmap. 

To illustrate how to do this, let's import pre-computed chromatin loops in `R`. 
These loops have been identified using `chromosight` 
(@Matthey_Doret_2020) on the contact matrix 
which we imported interactions from. 

```{r}
library(rtracklayer)
library(InteractionSet)
loops <- system.file('extdata', 'S288C-loops.bedpe', package = 'HiCExperiment') |> 
    import() |> 
    makeGInteractionsFromGRangesPairs()
loops
```

Similarly, borders have also been mapped with `chromosight`. We can also 
import them in `R`. 

```{r}
borders <- system.file('extdata', 'S288C-borders.bed', package = 'HiCExperiment') |> 
    import()
borders
```

Chromatin loops are stored in `GInteractions` while borders are `GRanges`. The 
former will be displayed as **off-diagonal circles** and the later as 
**on-diagonal diamonds** on the Hi-C heatmap. 

```{r}
plotMatrix(hic, loops = loops, borders = borders)
```

### Aggregated Hi-C maps

Finally, Hi-C map "snippets" (i.e. extracts) are often aggregated together to 
show an average signal. This analysis is sometimes referred to as APA (Aggregated
Plot Analysis). 

Aggregated Hi-C maps can be computed over a collection of `targets` using 
the `aggregate` function. These targets can be `GRanges` (to extract on-diagonal snippets) or `GInteractions` 
(to extract off-diagonal snippets). The `flankingBins` specifies how many 
matrix bins should be extracted on each side of the `targets` of interest. 

Here, we compute the aggregated Hi-C snippets of ± 15kb around each chromatin loop 
listed in `loops`. 

```{r}
hic <- zoom(hic, 1000)
aggr_loops <- aggregate(hic, targets = loops, flankingBins = 15)
aggr_loops
```

`aggregate` generates a `AggrHiCExperiment` object, a flavor of `HiCExperiment` 
class of objects. 

- `AggrHiCExperiment` objects have an extra `slices` slot. This 
stores a list of `array`s, one per `scores`. Each `array` is of 3 dimensions, 
`x` and `y` representing the heatmap axes, and `z` representing the index of the 
`target`. 
- `AggrHiCExperiment` objects also have a mandatory `topologicalFeatures` 
element named `targets`, storing the genomic loci provided in `aggregate`. 

```{r}
slices(aggr_loops)
dim(slices(aggr_loops, 'count'))
topologicalFeatures(aggr_loops, 'targets')
```

The resulting `AggrHiCExperiment` can be plotted using the same `plotMatrix` 
function with the arguments described above. 

```{r}
plotMatrix(
    aggr_loops, 
    use.scores = 'detrended', 
    scale = 'linear', 
    limits = c(-1, 1), 
    cmap = bgrColors()
)
```

# Session info {-}

::: {.callout-note collapse="true"}

## Click to expand 👇 {-}

```{r "session info"}
#| cache: false
sessioninfo::session_info(include_base = TRUE)
```

:::

# References {-}
