# Workflow 2: Chromosome compartment cohesion upon mitosis entry {-}

```{r}
#| echo: false
#| results: "hide"
#| message: false
#| warning: false
library(ggplot2)
library(cowplot)
library(purrr)
library(HiCExperiment)
library(HiContactsData)
library(fourDNData)
```

::: {.callout-note}

## Aims
This chapter illustrates how to: 

- Annotate compartments for a list of HiC experiments
- Generate saddle plots for a list of HiC experiments
- Quantify changes in interactions between compartments between different timepoints

:::

::: {.callout-important}
## Datasets

We leverage five chicken datasets in this notebook, published in @Gibcus_2018. 
They are all available from the 4DN data portal using the `fourDNData` package. 

- `4DNES9LEZXN7`: chicken cell culture blocked in G2 
- `4DNESNWWIFZU`: chicken cell culture released from G2 block (5min)
- `4DNESGDXKM2I`: chicken cell culture released from G2 block (10min)
- `4DNESIR416OW`: chicken cell culture released from G2 block (15min)
- `4DNESS8PTK6F`: chicken cell culture released from G2 block (30min)

:::

## Importing data

The 4DN consortium provides access to the datasets published in @Gibcus_2018. 
in `R`, they can be obtained thanks to the `fourDNData` gateway package. 

::: {.callout-warning}

### Beware
The first time the following chunk of code is executed, it will cache a large 
amount of data (mostly consisting of contact matrices stored in `.mcool` files).

:::

```{r}
library(HiCExperiment)
library(fourDNData)
library(BiocParallel)
samples <- list(
    '4DNES9LEZXN7' = 'G2 block', 
    '4DNESNWWIFZU' = 'prophase (5m)', 
    '4DNESGDXKM2I' = 'prophase (10m)', 
    '4DNESIR416OW' = 'prometaphase (15m)', 
    '4DNESS8PTK6F' = 'prometaphase (30m)' 
)
bpparam <- MulticoreParam(workers = 5, progressbar = TRUE)
hics <- bplapply(names(samples), fourDNHiCExperiment, BPPARAM = bpparam)
```

```{r}
names(hics) <- samples

hics[["G2 block"]]
```

## Plotting whole chromosome matrices

We can visualize the five different Hi-C maps on the entire chromosome `3` with 
`HiContacts` by iterating over each of the `HiCExperiment` objects. 

```{r}
library(purrr)
library(HiContacts)
library(ggplot2)
pl <- imap(hics, ~ .x['chr3'] |> 
    zoom(100000) |> 
    plotMatrix(use.scores = 'balanced', limits = c(-4, -1), caption = FALSE) + 
    ggtitle(.y)
)
library(cowplot)
plot_grid(plotlist = pl, nrow = 1)
```

This highlights the progressive remodeling of chromatin into condensed 
chromosomes, starting as soon as 5' after release from G2 phase. 

## Zooming on a chromosome section

Zooming on a chromosome section, we can plot the Hi-C autocorrelation matrix for each timepoint. These matrices are generally used to highlight the overall correlation of interaction profiles between different segments of a chromosome section (see [Chapter 5](matrix-centric.qmd#computing-autocorrelated-map) for more details). 

```{r}
## --- Format compartment positions of chr. 4 segment
.chr <- 'chr4'
.start <- 59000000L
.stop <- 75000000L
library(GenomicRanges)
coords <- GRanges(paste0(.chr, ':', .start, '-', .stop))
compts_df <- topologicalFeatures(hics[["G2 block"]], "compartments") |> 
    subsetByOverlaps(coords, type = 'within') |> 
    as.data.frame()
compts_gg <- geom_rect(
    data = compts_df, 
    mapping = aes(xmin = start, xmax = end, ymin = -500000, ymax = 0, alpha = compartment), 
    col = 'black', inherit.aes = FALSE
)

## --- Subset contact matrices to chr. 4 segment and computing autocorrelation scores
g2 <- hics[["G2 block"]] |> 
    zoom(100000) |> 
    subsetByOverlaps(coords) |>
    autocorrelate()
pro5 <- hics[["prophase (5m)"]] |> 
    zoom(100000) |> 
    subsetByOverlaps(coords) |>
    autocorrelate()
pro30 <- hics[["prometaphase (30m)"]] |> 
    zoom(100000) |> 
    subsetByOverlaps(coords) |>
    autocorrelate()

## --- Plot autocorrelation matrices
plot_grid(
    plotMatrix(
        subsetByOverlaps(g2, coords),
        use.scores = 'autocorrelated', 
        scale = 'linear', 
        limits = c(-1, 1), 
        cmap = bwrColors(), 
        maxDistance = 10000000, 
        caption = FALSE
    ) + ggtitle('G2') + compts_gg,
    plotMatrix(
        subsetByOverlaps(pro5, coords),
        use.scores = 'autocorrelated', 
        scale = 'linear', 
        limits = c(-1, 1), 
        cmap = bwrColors(), 
        maxDistance = 10000000, 
        caption = FALSE
    ) + ggtitle('Prophase 5min') + compts_gg,
    plotMatrix(
        subsetByOverlaps(pro30, coords),
        use.scores = 'autocorrelated', 
        scale = 'linear', 
        limits = c(-1, 1), 
        cmap = bwrColors(), 
        maxDistance = 10000000, 
        caption = FALSE
    ) + ggtitle('Prometaphase 30min') + compts_gg,
    nrow = 1
)
```

These correlation matrices suggest that there are two different regimes of chromatin compartment remodeling in this chromosome section: 

1. Correlation scores between genomic bins within the compartment A remain positive 5' after G2 release (albeit reduced compared to G2 block) and eventually become null 30' after G2 release. 
2. Correlation scores between genomic bins within the compartment B are overall null as soon as 5' after G2 release. 

## Generating saddle plots

Saddle plots are typically used to measure the `observed` vs. `expected` interaction scores within or between genomic loci belonging to A and B compartments. Here, they can be used to check whether the two regimes of chromatin compartment remodeling are observed genome-wide. 

Non-overlapping genomic windows are grouped by `nbins` quantiles (typically between 10 and 50 bins) according to their A/B compartment eigenvector value, from lowest eigenvector values (i.e. strongest B compartments) to highest eigenvector values (i.e. strongest A compartments). The average `observed` vs. `expected` interaction scores are computed for pairwise eigenvector quantiles and plotted in a 2D heatmap. 

```{r}
pl <- imap(hics, ~ plotSaddle(.x, nbins = 38, BPPARAM = bpparam) + ggtitle(.y)) 
plot_grid(plotlist = pl, nrow = 1)
```

These plots confirm the previous observation made on chr. `4` and reveal that intra-B compartment interactions are generally lost 5' after G2 release, while intra-A interactions take up to 15' after G2 release to disappear.

::: {.callout-warning}

### Beware
The `plotSaddle()` function requires an eigenvector corresponding to A/B 
compartments. In this example, this eigenvector is recovered from the 4DN data 
portal. If not already available, this eigenvector can be computed from the 
contact matrix using the `getCompartments()` function.

:::

## Quantifying interactions within and between compartments

We can leverage the replicate-merged contact matrices to quantify the 
interaction frequencies within A or B compartments or between A and B 
compartments, at different timepoints. 

We can use the A/B compartment annotations obtained at the `G2 block` timepoint and extract 
`O/E` (observed vs expected) scores for interactions within A or B 
compartments or between A and B compartments, at different timepoints. 

```{r}
## --- Extract the A/B compartments identified in G2 block
compts <- topologicalFeatures(hics[["G2 block"]], "compartments")
compts$ID <- paste0(compts$compartment, seq_along(compts))

## --- Iterate over timepoints to extract `detrended` (O/E) scores and 
##     compartment annotations
library(tibble)
library(plyranges)
df <- imap(hics[c(1, 2, 5)], ~ {
    ints <- cis(.x) |> ## Filter out trans interactions
        detrend() |> ## Compute O/E scores
        interactions() ## Recover interactions 
    ints$comp_first <- join_overlap_left(anchors(ints, "first"), compts)$ID
    ints$comp_second <- join_overlap_left(anchors(ints, "second"), compts)$ID
    tibble(
        sample = .y, 
        bin1 = ints$comp_first, 
        bin2 = ints$comp_second, 
        dist = InteractionSet::pairdist(ints), 
        OE = ints$detrended 
    ) |> 
        filter(dist > 5e6) |>
        mutate(type = dplyr::case_when(
            grepl('A', bin1) & grepl('A', bin2) ~ 'AA',
            grepl('B', bin1) & grepl('B', bin2) ~ 'BB',
            grepl('A', bin1) & grepl('B', bin2) ~ 'AB',
            grepl('B', bin1) & grepl('A', bin2) ~ 'BA'
        )) |> 
        filter(bin1 != bin2)
}) |> list_rbind() |> mutate(
    sample = factor(sample, names(hics)[c(1, 2, 5)])
)
```

We can now plot the changes in O/E scores for intra-A, intra-B, A-B 
or B-A interactions, splitting boxplots by timepoint. 

```{r}
ggplot(df, aes(x = type, y = OE, group = type, fill = type)) + 
    geom_boxplot(outlier.shape = NA) + 
    facet_grid(~sample) + 
    theme_bw() + 
    ylim(c(-2, 2))
```

This visualization suggests that interactions between genomic loci belonging to 
the B compartment are lost more rapidly than those between genomic loci belonging 
to the A compartment, when cells are released from G2 to enter mitosis. 

# Session info {-}

::: {.callout-note collapse="true"}

## Click to expand 👇 {-}

```{r "session info"}
#| cache: false
sessioninfo::session_info(include_base = TRUE)
```

:::

# References {-}
