# Workflow 3: Inter-centromere interactions in yeast {-}

::: {.callout-note collapse="true"}

## Pre-loading packages and objects 📦 {-}

```{r}
library(ggplot2)
library(purrr)
library(GenomicRanges)
library(InteractionSet)
library(HiCExperiment)
library(HiContactsData)
library(multiHiCcompare)
```

:::

::: {.callout-note}

## Aims
This chapter illustrates how to plot the aggregate signal over pairs of 
genomic ranges, in this case pairs of yeast centromeres. 

:::

::: {.callout-important}

## Datasets
We leverage two yeast datasets in this notebook. 

- One from a WT yeast strain in G1 phase
- One from a WT yeast strain in G2/M phase

:::

## Importing Hi-C data and plotting contact matrices

```{r}
library(HiContactsData)
library(HiContacts)
library(purrr)
library(ggplot2)
hics <- list(
    'G1' = import(HiContactsData('yeast_g1', 'mcool'), format = 'cool', resolution = 4000),
    'G2M' = import(HiContactsData('yeast_g2m', 'mcool'), format = 'cool', resolution = 4000)
)
imap(hics, ~ plotMatrix(
    .x, use.scores = 'balanced', limits = c(-4, -1), caption = FALSE
) + ggtitle(.y))
```

We can visually appreciate that inter-chromosomal interactions, notably 
between centromeres, are less prominent in G2/M.

## Checking P(s) and cis/trans interactions ratio

```{r}
library(dplyr)
pairs <- list(
    'G1' = PairsFile(HiContactsData('yeast_g1', 'pairs')),
    'G2M' = PairsFile(HiContactsData('yeast_g2m', 'pairs')) 
)
ps <- imap_dfr(pairs, ~ distanceLaw(.x, by_chr = TRUE) |> 
    mutate(sample = .y) 
)
plotPs(ps, aes(x = binned_distance, y = norm_p, group = interaction(sample, chr), color = sample)) + 
    scale_color_manual(values = c('black', 'red'))
plotPsSlope(ps, ggplot2::aes(x = binned_distance, y = slope, group = interaction(sample, chr), color = sample)) + 
    scale_color_manual(values = c('black', 'red'))
```

This confirms that interactions in cells synchronized in G2/M are enriched 
for 10-30kb-long interactions. 

```{r}
ratios <- imap_dfr(hics, ~ cisTransRatio(.x) |> mutate(sample = .y))
ggplot(ratios, aes(x = chr, y = trans_pct, fill = sample)) + 
    geom_col() + 
    labs(x = 'Chromosomes', y = "% of trans interactions") + 
    scale_y_continuous(labels = scales::percent) + 
    facet_grid(~sample)
```

We can also highlight that trans (inter-chromosomal) interactions are proportionally 
decreasing in G2/M-synchronized cells. 

## Centromere virtual 4C profiles

```{r}
data(centros_yeast)
v4c_centro <- imap_dfr(hics, ~ virtual4C(.x, GenomicRanges::resize(centros_yeast[2], 8000)) |> 
    as_tibble() |> 
    mutate(sample = .y) |> 
    filter(seqnames == 'IV')
) 
ggplot(v4c_centro, aes(x = start, y = score, fill = sample)) +
    geom_area() +
    theme_bw() +
    labs(
        x = "chrIV position", 
        y = "Contacts with chrII centromere", 
        title = "Interaction profile of chrII centromere"
    ) + 
    coord_cartesian(ylim = c(0, 0.015))
```

## Aggregated 2D signal over all pairs of centromeres

We can start by computing all possible pairs of centromeres. 

```{r}
centros_pairs <- lapply(1:length(centros_yeast), function(i) {
    lapply(1:length(centros_yeast), function(j) {
        S4Vectors::Pairs(centros_yeast[i], centros_yeast[j])
    })
}) |> 
    do.call(c, args = _) |>
    do.call(c, args = _) |> 
    InteractionSet::makeGInteractionsFromGRangesPairs()
centros_pairs <- centros_pairs[anchors(centros_pairs, 'first') != anchors(centros_pairs, 'second')]

centros_pairs
```

Then we can aggregate the Hi-C signal over each pair of centromeres.

```{r}
aggr_maps <- purrr::imap(hics, ~ {
    aggr <- aggregate(.x, centros_pairs, maxDistance = 1e999)
    plotMatrix(
        aggr, use.scores = 'balanced', limits = c(-5, -1), 
        cmap = HiContacts::rainbowColors(), 
        caption = FALSE
    ) + ggtitle(.y)
})

cowplot::plot_grid(plotlist = aggr_maps, nrow = 1)
```

## Aggregated 1D interaction profile of centromeres 

One can generalize the previous virtual 4C plot, by extracting the interaction profile 
between all possible pairs of centromeres in each dataset. 

```{r}
df <- map_dfr(1:{length(centros_yeast)-1}, function(i) {
    centro1 <- GenomicRanges::resize(centros_yeast[i], fix = 'center', 8000)
    map_dfr({i+1}:length(centros_yeast), function(j) {
        centro2 <- GenomicRanges::resize(centros_yeast[j], fix = 'center', 80000)
        gi <- InteractionSet::GInteractions(centro1, centro2)
        imap_dfr(hics, ~ .x[gi] |> 
            interactions() |> 
            as_tibble() |>
            mutate(
                sample = .y, 
                center = center2 - start(GenomicRanges::resize(centro2, fix = 'center', 1))
            ) |> 
            select(sample, seqnames1, seqnames2, center, balanced)
        )
    })
}) 
ggplot(df, aes(x = center/1e3, y = balanced)) + 
    geom_line(aes(group = interaction(seqnames1, seqnames2)), alpha = 0.03, col = "black") + 
    geom_smooth(col = "red", fill = "red") + 
    theme_bw() + 
    theme(legend.position = 'none') + 
    labs(
        x = "Distance from centromere (kb)", y = "Normalized interaction frequency", 
        title = "Centromere pairwise interaction profiles"
    ) +
    facet_grid(~sample)
```

# Session info {-}

::: {.callout-note collapse="true"}

## Click to expand 👇 {-}

```{r "session info"}
#| cache: false
sessioninfo::session_info(include_base = TRUE)
```

:::

# References {-}
